Review Paper Draft

### 2020-08-13 main.tex
Sections 4 and 5: small edits, notation fixes

Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^\T + \sigmaee \bU \bU^\T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^\T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^\T\\
    &=\bU \bW^2 \bU^\T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^\T \bX$ and $\byT = \bU^\T \by$, where $\bW = \text{diag}\{w_1, ... w_n\}$ and $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
In the presence of an intercept and unpenalized covariates, \eqref{eqn:rtY_distn} becomes 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N\left(\bV^{-\frac{1}{2}} \bX{^\prime} \bbeta{^\prime}, \bI \right)
\end{equation}
where $\bX{^\prime} = \begin{bmatrix} \bX_0 & \bX \end{bmatrix}$ and $\bbeta{^\prime} = \begin{bmatrix} \bbeta_0^\T & \bbeta^\T \end{bmatrix}^\T$. This in turn leads to a new version of \eqref{eqn:lmm_obj}, our objective function,  
\begin{equation}
Q(\bbeta | \bX{^\prime}, \byT, \bW) = || \bW (\byT - \bXT{^\prime} \bbeta{^\prime})||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\bXT{^\prime} = \bU^\T \bX{^\prime}$, so that all of the data are transformed, but only the elements of $\bbeta{^\prime}$ corresponing to SNPs are penalized.

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \citep{Rakitsch2012} uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix \citep{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude is inconsistent with the confounding environmental effects. This has the effect of ``washing out'' bias in $\hat{\bbeta}$, which is of interest, while inflating the variance. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg I_p)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}
\anna{beta-hat and eta-hat seems like a bug with ggplot2 update (?) - checking on this}

@@

In addition, we have found that LMM-Lasso-ggmix performs poorly in a low-dimensional setting due to the ill-conditioning of the RRM. Given its intended use for the analysis of high-dimensional data, this fact could be of little consequence to practitioners. However, although its estimation accuracy in high dimensions is on par with that of LMM-Lasso-Rakitsch, as shown in figure \ref{fig:eta_beta_mse}, this is also a setting in which the computational time required may be prohibitive for practical applications. Figure \ref{fig:time} shows the ratio of computing time required for LMM-Lasso-ggmix compared to that of LMM-Lasso-Rakitsch, for various sample sizes. The horizontal dashed line at 1 indicates both methods require the same computing time. Despite its comparable performance in accurately estimating $\hat{\bbeta}$, LMM-Lasso-ggmix takes requires an average of \anna{73.4} times the computing time of LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method than LMM-Lasso-ggmix.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion} \label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this work, we have reviewed the concept of population stratification and explicitly defined the environmental (i.e., non-genetic) mechanism by which it may confound the genotype-phenotype relationship. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso.

We have detailed the assumptions made by PC-Lasso and LMM-Lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which simulation schemes to evaluate these methods may fail to accurately capture the effects of environmental heterogeneity. In turn, we propose and utilize a simulation approach in which the estimation bias introduced by environmental heterogeneity may be evaluated. We then compare the estimation accuracy of a naive lasso, PC-Lasso, and LMM-Lasso in various confounding scenarios.

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the naive lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the naive lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso when (1) fine subpopulation structure is present and (2) when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from instability in low dimensions. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably in high dimensions, LMM-Lasso-ggmix requires an average of 73.4 and as much as 167.57 \anna{73.4 is the mean ratio + sd, but the max absolute ratio is 167.57 - which should I report?} times the computing time of LMM-Lasso-Rakitsch due to its iterative estimation procedure. Thirdly, while the additional time required of LMM-Lasso-ggmix may be worthwhile if it yielded more accurate estimates of $\eta$, or narrow-sense heritability, we have been unable to identify a scenario where its narrow-sense heritability estimates are greatly superior to those of LMM-Lasso-Rakitsch. We have implemented the LMM-Lasso-Rakitsch method using the equivalent, but more interpretable, $\eta$ paramaterization in the \texttt{R} package, \texttt{penalizedLMM (final name tbd)}, available on the author's github page: \url{https://github.com/areisett/penalizedLMM}.

@@

It is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs may be an attractive avenue for future research.

>>>>

Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^\T + \sigmaee \bU \bU^\T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^\T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^\T\\
    &=\bU \bW^2 \bU^\T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^\T \bX$ and $\byT = \bU^\T \by$, where $\bW = \text{diag}\{w_1, ... w_n\}$ and $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
In the presence of an intercept and unpenalized covariates, \eqref{eqn:rtY_distn} becomes 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N\left(\bV^{-\frac{1}{2}} \bX{^\prime} \bbeta{^\prime}, \bI \right)
\end{equation}
where $\bX{^\prime} = \begin{bmatrix} \bX_0 & \bX \end{bmatrix}$ and $\bbeta{^\prime} = \begin{bmatrix} \bbeta_0^\T & \bbeta^\T \end{bmatrix}^\T$. This in turn leads to a new version of \eqref{eqn:lmm_obj}, our objective function,  
\begin{equation}
Q(\bbeta | \bXT{^\prime}, \byT, \bW) = || \bW (\byT - \bXT{^\prime} \bbeta{^\prime})||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\bXT{^\prime} = \bU^\T \bX{^\prime}$, so that all of the data are transformed, but only the elements of $\bbeta{^\prime}$ corresponing to SNPs are penalized.

@@

We have so far used the term $\bW$ to describe the LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \citep{Rakitsch2012} uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix \citep{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude is inconsistent with the confounding environmental effects. This has the effect of ``washing out'' bias in $\bbetaHat$, which is of interest, while inflating the variance. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg I_p)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

In addition, we have found that LMM-Lasso-ggmix performs poorly in a low-dimensional setting due to the ill-conditioning of the RRM. Given its intended use for the analysis of high-dimensional data, this fact could be of little consequence to practitioners. However, although its estimation accuracy in high dimensions is on par with that of LMM-Lasso-Rakitsch, as shown in figure \ref{fig:eta_beta_mse}, this is also a setting in which the computational time required may be prohibitive for practical applications. \anna{time interpretation depends on reason ggmix is slow...} Figure \ref{fig:time} shows the ratio of computing time required for LMM-Lasso-ggmix compared to that of LMM-Lasso-Rakitsch, for various sample sizes. The horizontal dashed line at 1 indicates both methods require the same computing time. Despite its comparable performance in accurately estimating $\bbetaHat$, LMM-Lasso-ggmix takes requires an average of \anna{73.4} times the computing time of LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method than LMM-Lasso-ggmix.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion} \label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this work, we have reviewed the concept of population stratification and explicitly defined the environmental mechanism by which it may confound the genotype-phenotype relationship. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso.

We have detailed the assumptions made by PC-Lasso and LMM-Lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which simulation schemes to evaluate these methods may fail to accurately capture the effects of environmental heterogeneity. In turn, we propose and utilize a simulation approach in which the estimation bias introduced by environmental heterogeneity may be evaluated. We then compare the estimation accuracy of a \anna{naive} Lasso, PC-Lasso, and LMM-Lasso in various confounding scenarios.

Unsurprisingly, we find that PC-Lasso and LMM-Lasso outperform the \anna{naive} Lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-Lasso may hinder estimation accuracy and perform more poorly than even the \anna{naive} Lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-Lasso outperforms PC-Lasso when (1) fine subpopulation structure is present and (2) when confounding effects are nonlinear and discordant in nature. In the presence of coarse subpopulation structure which corresponds to concordant, dichotomous environmental effects, PC-Lasso may outperform LMM-Lasso in estimation accuracy. However, LMM-Lasso is more robust to subpopulation structure as well as nonlinear and discordant environmental effects. Therefore, LMM-Lasso is a more robust method for practitioners seeking to obtain unbiased estimates when the exact nature of the population structure and corresponding environmental heterogeneity are unknown.

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from instability in low dimensions. \anna{depends on ggmix time results} Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably in high dimensions, LMM-Lasso-ggmix requires an average of 73.4 and as much as 167.57 \anna{73.4 is the mean ratio + sd, but the max absolute ratio is 167.57 - which should I report?} times the computing time of LMM-Lasso-Rakitsch due to its iterative estimation procedure. Thirdly, while the additional time required of LMM-Lasso-ggmix may be worthwhile if it yielded more accurate estimates of $\eta$, or narrow-sense heritability, we have been unable to identify a scenario where its narrow-sense heritability estimates are greatly superior to those of LMM-Lasso-Rakitsch. We have implemented the LMM-Lasso-Rakitsch method using the equivalent, but more interpretable, $\eta$ paramaterization in the \texttt{R} package, \texttt{penalizedLMM (final name tbd)}, available on the author's github page: \url{https://github.com/areisett/penalizedLMM}.

@@

It is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs are an attractive avenue for future research.
